Dimensionality reduction techniques such as PCA, UMAP \cite{mcinnes2018umap}, and t-SNE \cite{van2008visualizing} are widely used for visualizing high-dimensional data. While PCA preserves global linear structure, nonlinear methods like UMAP and t-SNE focus on local neighborhood preservation, often at the expense of global relationships.

PHATE \cite{moon2019visualizing} was introduced to better capture both local and global structure through diffusion geometry, and has been successfully applied in biological and scientific data analysis. However, its application to large-scale textual data remains less explored.

In hierarchical clustering evaluation, traditional metrics such as ARI and AMI do not capture structural relationships between clusters. Prior work has introduced structure-aware metrics such as LCA-F1 \cite{kosmopoulos2015} and dendrogram purity \cite{heller2005bayesian}, which evaluate agreement between hierarchical structures. These metrics have been shown to provide more meaningful evaluation for hierarchical classification tasks \cite{kosmopoulos2015}

\subsection{Dimensionality reduction \& visualization}
To evaluate its utility for text, we compare PHATE visualizations against those from PCA, UMAP, and t-SNE.
\begin{description}
    \item[PCA] reduces dimensionality by projecting data onto the leading eigenvectors of the covariance matrix, preserving global linear structure.
    \item[UMAP] uses a weighted k-nearest neighbor (kNN) graph and a nonlinear embedding based on force-directed layout \cite{mcinnes2018umap}.
    \item[t-SNE] forms a probability distribution over pairwise dissimilarities and embeds points such that the KL divergence between the original and embedded distributions is minimized \cite{van2008visualizing}
    \item[PHATE] also constructs a weighted kNN graph, row-normalizes its adjacency matrix to a Markov operator, amplifies global structure by powering the operator, computes information distances using row-wise log-transforms, and applies metric multidimensional scaling to derive low-dimensional coordinates \cite{moon2019visualizing}.
    \item[PaCMAP] (Pairwise Controlled Manifold Approximation) balances preservation of both local and global structure by explicitly controlling pairwise relationships during embedding optimization \cite{wang2021pacmap}.
    \item[TriMAP] emphasizes preservation of global structure by optimizing triplet relationships between points, ensuring that relative distances are maintained in the low-dimensional embeddingTriMAP \cite{trimap}.
\end{description}
\textit{Qualitative comparisons} are made using datasets with ground-truth metadata. Evaluation criteria include cluster coherence, separation of known categories, and preservation of trajectories or arcs in embedding space.
\subsection{Text Embeddings}

We generate dense semantic embeddings using two models: all-MiniLM-L6-v2 and Qwen3-Embedding-0.6B. Each document is converted into a high-dimensional vector representation capturing semantic meaning.

For Qwen embeddings, text inputs are tokenized and transformed into 1024-dimensional vectors. Mean pooling is applied to obtain a single representation per document, followed by L2 normalization to ensure consistent similarity comparisons.
\subsection{Hierarchical clustering}

To assess PHATE in clustering, we:
\begin{enumerate}
\item Generate datasets with known hierarchical structure (see "Datasets").
\item Embed the data using pre-trained language models, including \texttt{all-MiniLM-L6-v2} and \texttt{Qwen3-Embedding-0.6B}.
\item Apply PCA, UMAP, PHATE, PaCMAP, and TriMAP for dimensionality reduction.
\item Cluster with agglomerative clustering (Ward's method), HDBSCAN, Diffusion Condensation, and HERCULES.
\end{enumerate}

\paragraph{Clustering Methods.}
We evaluate multiple clustering approaches on the reduced embeddings.

Agglomerative clustering constructs a hierarchical tree using Ward linkage, enabling direct evaluation of hierarchical structure.

HDBSCAN is a density-based clustering method that identifies clusters of varying density while labeling noise points automatically.

Diffusion Condensation builds hierarchical structure through an iterative diffusion process, progressively merging points based on similarity.

HERCULES is an LLM-based hierarchical clustering framework that constructs topic hierarchies using semantic reasoning and recursive partitioning.
These methods were selected to compare hierarchical and density-based clustering approaches under a unified embedding and reduction pipeline.
\subsection{Experimental Pipeline}

We apply dimensionality reduction methods, including PCA, UMAP, PHATE, PaCMAP, and TriMAP, to obtain low-dimensional representations.
Clustering is performed on each reduced representation using Agglomerative Clustering, HDBSCAN, Diffusion Condensation, and HERCULES. All methods operate on identical precomputed embeddings to ensure that performance differences arise solely from the dimensionality reduction and clustering methods.
\paragraph{Evaluation Metrics.}
We evaluate clustering performance using AMI, ARI, Fowlkes–Mallows (FM), and Rand Index. 
For dimensionality reduction, we use trustworthiness, continuity, Shepard diagrams, Spearman correlation, and DEMaP.
\subsection{Datasets}
To assess the visual quality of PHATE embeddings, public comments from the Environmental Protection Agency (EPA) were sourced from the Federal Register via Regulations.gov as part of a related project studying public perceptions and preferences as related to environmental policy. The data was collected from an open source initiative called Mirrulations\footnote{https://registry.opendata.aws/mirrulations/}, which mirrors this public data through an Amazon Web Services S3 bucket. Once the data was gathered, the text was cleaned by removing HTML tags, special characters, URLs, stop words, and excessively short entries, focusing only on comments submitted after the year 2000. To handle long comments and improve compatibility with transformer models, a head-tail truncation strategy \cite{sun2019fine} was used to preserve meaningful content from the beginning and end of each comment. The cleaned data was embedded using the sentence-transformers model all-\texttt{MiniLM-L6-v2}. 
To evaluate clustering quality, we required hierarchical text datasets with known ground truth. As no available datasets met our application needs, we generated synthetic datasets using an LLM. We curated two thematic seeds—\textit{Energy, Ecosystems, and Humans} and \textit{Offshore Energy Impacts on Fisheries}—to span a range from general to domain-specific topics related to our parallel applied research in social-ecological systems modeling. From each seed, we recursively generated hierarchical structures of varying depth and breadth, simulating both shallow and deeply nested trees.

To mimic the irregularities of real-world ontologies, we varied the number of children per node at random. A labeling heuristic propagated parent labels to children while excluding the parents themselves from deeper levels, reflecting semantic containment without conflating hierarchical roles. To further simulate real-world messiness, we injected structural noise in the form of concepts from overlapping subtopics and ambiguous nodes. 
% These elements challenged models to infer latent structure and semantic grouping.
The resulting datasets were used for both quantitative benchmarking and qualitative analysis of PHATE’s capacity to capture semantic and hierarchical organization in text.

To assess performance across diverse hierarchical domains, we applied our methods to benchmark datasets with known hierarchies, including Web of Science paper keywords \cite{kowsari2017HDLTex}, Wikipedia article categories \cite{brummer2016dbpedia, NIPS2015_250cf8b5, yang2019xlnet}, Amazon product reviews \cite{yury_kashnitsky_2020}, arXiv \cite{arxiv_kaggle}, with hierarchical labels derived from the arXiv category taxonomy \cite{arxiv_taxonomy}. The data was accessed via a publicly available Kaggle mirror. We also use the RCV1 dataset \cite{lewis2004rcv1}.

\begin{table}[ht]
    \centering
    \caption{Number of datapoints in each dataset.}
    \begin{tabular}{l l l r}
    \hline
    \textbf{Category} & \textbf{Topic} & \textbf{Depth} & \textbf{N} \\
    \hline
    \multirow{4}{*}{Generated} 
     & \multirow{2}{*}{Ecosystems} & Deep & 19{,}469 \\
     &  & Shallow & 4{,}848 \\
     & \multirow{2}{*}{Fisheries} & Deep & 19{,}806 \\
     &  & Shallow & 6{,}708 \\
    \hline
    \multirow{5}{*}{Benchmark}
     & \multicolumn{2}{l}{WoS} & 46{,}985 \\
     & \multicolumn{2}{l}{DBpedia} & 60{,}794 \\
     & \multicolumn{2}{l}{Amazon} & 14{,}824 \\
     & \multicolumn{2}{l}{arXiv} & 30{,}000 \\
     & \multicolumn{2}{l}{RCV1} & 1{,}628 \\
    \hline
    \end{tabular}
    \label{tab:dataset_counts}
    \end{table}